Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-25 paper.qmd
Updated Figure 18 aspect ratio
changed to out-width

error_plot_one_curvy_hist_selected <- ggplot(error_df_one_curvy_abs, aes(x = sqrt_row_wise_total_error, y = density,
                                                                         colour = factor(error_cat_n,
                                                                                         levels = c("high", "low")))) +
  geom_point() +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  xlab(expression(e[hj])) +
  ylab("") +
  interior_annotation("b1",
                      position = c(0.95, 0.9),
                      cex = 1.5) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 15)
  )

error_plot_one_curvy_hist <- ggplot(error_df_one_curvy_abs,
                                    aes(x = sqrt_row_wise_total_error, y = density,
                                        colour = factor(error_cat_n2,
                                                        levels = c("high", "low")))) +
  geom_point() +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  xlab(expression(e[hj])) +
  ylab("") +
  interior_annotation("a1",
                      position = c(0.95, 0.9),
                      cex = 1.5) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 15)
  )
```

```{r}
#| label: link-model-one-dens-clust
plot_tsne_dens <- error_df_one_curvy_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = factor(error_cat_n2,
                             levels = c("high", "low")))) +
  geom_point(alpha=0.7) +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  theme(
    aspect.ratio = 1
  ) +
  interior_annotation("a2",
                      position = c(0.08, 0.9),
                      cex = 1.5)

plot_tsne_dens_selected <- error_df_one_curvy_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = factor(error_cat_n,
                             levels = c("high", "low")))) +
  geom_point(alpha=0.7) +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  theme(
    aspect.ratio = 1
  ) +
  interior_annotation("b2",
                      position = c(0.08, 0.9),
                      cex = 1.5)

@@

five_c_shaped_proj_tsne_model1 <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df, 
    axes = axes2, 
    circle = circle, 
    point_param = c(1.5, 0.5), # size, alpha, color
    line_param = c(0.8, 0.4, "#000000"), # linewidth, alpha
    plot_limits = c(-0.35, 0.35), 
    title = "a3", 
    cex = 1.5,
    position = c(0.08, 0.9),
    axis_text_size = 4, 
    is_category = TRUE) +
    scale_color_manual(values = c('#800026', '#d9d9d9')) +
    theme(
        legend.position = "none"
    )

@@

five_c_shaped_proj_tsne_model1_selected <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df, 
    axes = axes2, 
    circle = circle, 
    point_param = c(1.5, 0.5), # size, alpha, color
    line_param = c(0.8, 0.4, "#000000"), # linewidth, alpha
    plot_limits = c(-0.35, 0.35), 
    title = "a3", 
    cex = 1.5,
    position = c(0.08, 0.9),
    axis_text_size = 4, 
    is_category = TRUE) +
    scale_color_manual(values = c('#800026', '#d9d9d9')) +
    theme(
        legend.position = "none"
    )
```

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-height: 100%
#| fig-cap: "The sparse end of the structure shows a high model error, while the dense end shows a low model error. The tSNE layout of the C-shaped structure, with dense points (a1-a3) are colored according to their $4\\text{-}D$ model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high $4\\text{-}D$ error points in \\gD{} and $4\\text{-}D$. It helps in identifying the high $4\\text{-}D$ model errors that occur due to the sparse end. Videos of the linked plots are available at <>."

@@

<!--add langevitour screenshots and youtube animation link-->
```{r}
#| label: fig-mnist-tri-proj
#| fig-pos: H
#| fig-height: 4
#| fig-width: 12
#| out-height: 100%
#| fig-cap: "\\gD{} wireframe layout shows that there is a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster. The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a non-linear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). Video of the langevitour animation is available at <>."

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-height: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster. Videos of the linked plots are available at <>."

@@

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "The $10\\text{-}D$ model error in \\gD{} layout of the MNIST digit 1 data shows a pattern. Most low model errors are distributed along the big non-linear cluster, while most large model errors are distributed along the small cluster. The images associated with large model errors shows different patterns of digit 1, some inside (e) the non-linear structure and others outside (d). Along the non-linear cluster, the angle of digit 1 changes (a-c)."
#| fig-pos: H
#| fig-width: 10
#| fig-height: 7
#| out-width: 90%

>>>>

error_plot_one_curvy_hist_selected <- ggplot(error_df_one_curvy_abs, aes(x = sqrt_row_wise_total_error, y = density,
                                                                         colour = factor(error_cat_n,
                                                                                         levels = c("high", "low")))) +
  geom_point() +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  xlab(expression(e[hj])) +
  ylab("") +
  interior_annotation("b1",
                      position = c(0.95, 0.9),
                      cex = 1.5) +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

error_plot_one_curvy_hist <- ggplot(error_df_one_curvy_abs,
                                    aes(x = sqrt_row_wise_total_error, y = density,
                                        colour = factor(error_cat_n2,
                                                        levels = c("high", "low")))) +
  geom_point() +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  xlab(expression(e[hj])) +
  ylab("") +
  interior_annotation("a1",
                      position = c(0.95, 0.9),
                      cex = 1.5) +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))
```

```{r}
#| label: link-model-one-dens-clust
plot_tsne_dens <- error_df_one_curvy_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = factor(error_cat_n2,
                             levels = c("high", "low")))) +
  geom_point(alpha=0.7) +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  theme(aspect.ratio = 1) +
  interior_annotation("a2",
                      position = c(0.08, 0.9),
                      cex = 1.5)

plot_tsne_dens_selected <- error_df_one_curvy_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = factor(error_cat_n,
                             levels = c("high", "low")))) +
  geom_point(alpha=0.7) +
  scale_color_manual(values=c('#800026', '#d9d9d9')) +
  theme(aspect.ratio = 1) +
  interior_annotation("b2",
                      position = c(0.08, 0.9),
                      cex = 1.5)

@@

five_c_shaped_proj_tsne_model1 <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df, 
    axes = axes2, 
    circle = circle, 
    point_param = c(1.5, 0.5), # size, alpha, color
    line_param = c(0.8, 0.4, "#000000"), # linewidth, alpha
    plot_limits = c(-0.35, 0.35), 
    title = "a3", 
    cex = 1.5,
    position = c(0.08, 0.9),
    axis_text_size = 4, 
    is_category = TRUE) +
    scale_color_manual(values = c('#800026', '#d9d9d9')) +
    theme(aspect.ratio = 1,
          legend.position = "none")

@@

five_c_shaped_proj_tsne_model1_selected <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df, 
    axes = axes2, 
    circle = circle, 
    point_param = c(1.5, 0.5), # size, alpha, color
    line_param = c(0.8, 0.4, "#000000"), # linewidth, alpha
    plot_limits = c(-0.35, 0.35), 
    title = "a3", 
    cex = 1.5,
    position = c(0.08, 0.9),
    axis_text_size = 4, 
    is_category = TRUE) +
    scale_color_manual(values = c('#800026', '#d9d9d9')) +
    theme(aspect.ratio = 1,
          legend.position = "none")
```

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "The sparse end of the structure shows a high model error, while the dense end shows a low model error. The tSNE layout of the C-shaped structure, with dense points (a1-a3) are colored according to their $4\\text{-}D$ model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high $4\\text{-}D$ error points in \\gD{} and $4\\text{-}D$. It helps in identifying the high $4\\text{-}D$ model errors that occur due to the sparse end. Videos of the linked plots are available at <>."

@@

<!--add langevitour screenshots and youtube animation link-->
```{r}
#| label: fig-mnist-tri-proj
#| fig-pos: H
#| fig-height: 4
#| fig-width: 12
#| out-width: 100%
#| fig-cap: "\\gD{} wireframe layout shows that there is a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster. The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a non-linear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). Video of the langevitour animation is available at <>."

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-width: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big non-linear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster. Videos of the linked plots are available at <>."

@@

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "The $10\\text{-}D$ model error in \\gD{} layout of the MNIST digit 1 data shows a pattern. Most low model errors are distributed along the big non-linear cluster, while most large model errors are distributed along the small cluster. The images associated with large model errors shows different patterns of digit 1, some inside (e) the non-linear structure and others outside (d). Along the non-linear cluster, the angle of digit 1 changes (a-c)."
#| fig-pos: H
#| fig-width: 7
#| fig-height: 5
#| out-width: 90%
